---
title: "GMQL: GenoMetrics Query Language"
author: "Simone pallotta"
date: "`r Sys.Date()`"
output: BiocStyle::pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

vaffanculo

# Introduction

Improvement of sequencing technologies and data processing pipelines is rapidly providing sequencing data, with associated high-level features, of many individual genomes in multiple biological and clinical conditions.
For this purpose GMQL has been proposed a high-level, declarative GenoMetric Query Language (GMQL) and a toolkit for its use.

## Purpose

This package provides a set of functions to create, manipulate and extract genomic data from different datasources from local and remote datasets.

These functios allow performing complex queries without the knowledge of GMQL syntax

# Dataset

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::pdf_document:
        css: mystyles.css

# Environments

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::pdf_document:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## Local

## Global

# Local Environment 

## Datasource

## Queries

## Execution

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

# More Example


